\message{ !name(thesis.tex)}\documentclass[12pt, a4paper, twoside]{report}

\usepackage{setspace}
\onehalfspacing
% \doublespacing
\usepackage{amsmath, amssymb, amsthm, mathtools,mathrsfs}
\allowdisplaybreaks % to break align maths over pages
\usepackage[colorlinks=true, linkcolor=blue, citecolor=blue]{hyperref}
\usepackage{bbm, bm}
\usepackage{url}
\usepackage{fancyhdr} %, xspace, psfrag, setspace, supertabular
\usepackage{algorithm, algpseudocode}
\usepackage{graphicx}
\usepackage{geometry} % min OU requirements [inner=40mm, outer=15mm, top=15mm, bottom=20mm]
\usepackage[dvipsnames]{xcolor}
\usepackage{tikz}
\usetikzlibrary{automata,positioning,decorations.pathreplacing}
\tikzset{>=latex}
\usepackage[procnames]{listings} % Allows inclusion (and syntax highlighting) of C code blocks
\usepackage[font={small}]{caption}
\usepackage{subcaption}
\usepackage{footnote} 
\makesavenoteenv{tabular} % to be able to use footnotes in tabulars
\makesavenoteenv{table} % to be able to use footnotes in tables
\usepackage{stmaryrd} % for averaging operator
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{mathdots}


\begin{document}

\message{ !name(thesis.tex) !offset(-3) }




\input{defs}
\input{decors}
\input{title}

\setcounter{secnumdepth}{-2}% default for "report" is 2
\chapter{Abstract}
Enumerating permutations is sometimes hard and sometimes tedious.\\
\noindent Packing permutations is often hard and always tedious.

\chapter{Acknowledgements}
I would like to thank my supervisor Robert Brignall for finding the balance between guidance and freedom. I especially value that we had fun doing mathematics.\\

My acknowledgements go to the Department of Mathematics and Statistics at The Open University for funding my PhD and providing support in various forms throughout the past three years.\\

Special thanks go to my fellow students at The Open University who made my time there enjoyable.
\setcounter{secnumdepth}{2}

\tableofcontents

\chapter{Enumeration}
%\input{juxt}

\section{Iterated juxtapositions}
The goal of this section is to juxtapose a context-free permutation class $\C$ with a finite number of monotone classes $\M_1|\ldots|\M_k$. To begin, we assume that $\M_i$ are monotone increasing permutation classes. We will later argue that for decreasing classes, a symmetry of our argument applies. Our approach to juxtapositions in this section generalises the work in Section~\ref{sec:catalanjuxt} in two directions. One, the condition that $\C$ is a Catalan permutation class is replaced by requiring $\C$ to only be context-free. This is a significant relaxation. Two, juxtaposition from the right is iterated a finite number of times instead of just once. The trade-off comes in the form of a weaker result. We do not enumerate these grid classes explicitly. Instead, we prove that their generating functions are well-behaved, algebraic in particular. Before we proceed with the statement of the result, let us set the scene. The following definition is taken from Flajolet and Sedgewick~\cite{analcomb}, Section I.5.4, Definition I.13.

\begin{definition}[Context-free specification]
  A class $\C$ is said to be \emph{context-free} if it coincides with the first component $\S_1$ of a system of equations
  \begin{align}
  \begin{aligned}
    \begin{cases}
      \S_1 &= f_1(\Z,\S_1,\ldots,\S_r)\\
      &\vdots\\
      \S_r &= f_r(\Z,\S_1,\ldots,\S_r)
    \end{cases}
  \end{aligned}
  \end{align}
  where each $f_i$ is a constructor that only involves operations of combinatorial sum ($+$) and cartesian product ($\times$), as well as the neutral/empty class $\EE = \{\emptyset\}$.
\end{definition}

\begin{definition}[Tracking the rightmost point]
  We say that a context-free class specification of $\C$ \emph{tracks the rightmost point by its vertical position} if it is combinatorialy isomorphic to the context-free class with the specification $\rS$ below and all cartesian products in $\rS$ are left-to-right as they occur bottom-to-top in $\C$. The asterisk (${}^*$) in $\rS$ marks the rightmost point or the block which contains the rightmost point inside $\C_i^*$.
  \begin{align*}
\rS &=  \begin{rcases}\begin{dcases}
  \C_0^* &= f_0^*(\Z,\Z^*,\C_0,\ldots,\C_r,\C_0^*,\ldots,\C_r^*)\\
       &\vdots\\
  \C_r^* &= f_r^*(\Z,\Z^*,\C_0,\ldots,\C_r,\C_0^*,\ldots,\C_r^*)\\
  \C_0 &= f_0(\Z,\C_0,\ldots,\C_r)\\
       &\vdots\\
       \C_r &= f_r(\Z,\C_0,\ldots,\C_r)
       \end{dcases}
     \end{rcases}
   \end{align*}
\end{definition}
If a class $\C^*$ tracks the rightmost point as outlined above, we refer to $\C^*$ as a \emph{starred class}. Similarly, $\Z^*$ is a \emph{starred point}. On the other hand, every other class or point is \emph{starless class} or \emph{starless point}. For example, $\C$ and $\Z$. Given the above definition, we assume from now on that every permutation class $\C$ is context-free and admits a specification that tracks the rightmost point. 
\begin{example}
  The following is a context-free specification for the class of separable permutations (non-empty):
  \begin{align*}
    \S &= \Z + \S_{\oplus}\S + \S_{\ominus}\S\\
    \S_{\ominus} &= \Z + \S_\oplus\S\\
    \S_\oplus &= \Z + \S_\ominus\S,
  \end{align*}
  where $\S_\ominus$ and $\S_\oplus$ stand for skew-indecomposable and sum-indecomposable permutation classes. Now, among all context-free specifications of $\S$, we insist on picking the following one:
  \begin{align}
    \begin{aligned}
    \S^* &= \Z^* + \S_{\oplus}\S^* + \S^*\S_{\ominus}\\
    \S &= \Z + \S_{\oplus}\S + \S\S_{\ominus}\\
    \S_{\ominus} &= \Z + \S_\oplus\S\\
    \S_\oplus &= \Z + \S\S_\ominus.
    \label{eq:sepstarcs}
  \end{aligned}
  \end{align}
  In~\eqref{eq:sepstarcs}, the class of separable permutations is represented according to the decomposition in~\eqref{eq:sepstar}, encoding vertical order as left-to-right.
  \begin{align}
    \begin{aligned}
      \S^* &= \Z + \cplusc{\S_\oplus}{\S^*} + \cminusc{\S^*}{\S_\ominus}\\
      \S &= \Z + \cplusc{\S_\oplus}{\S} + \cminusc{\S}{\S_\ominus}\\
      \S_\ominus &= \Z + \cplusc{\S_\oplus}{\S}\\
      \S_\oplus &= \Z + \cminusc{\S}{\S_\ominus}
    \label{eq:sepstar}
  \end{aligned}
  \end{align}
\end{example}
Notice that requiring the cartesian product to be non-commutative (keeping track of vertical positions) does not interfere with the combinatorial specification being context-free. Indeed, we merely pick a specific combinatorial specification out of all context-free ones for a given class $\C$. We later use these properties later to decorate the combinatorial specification with monotone sequences on the right-hand side. Let us give one more example how vertical order translates into left-to-right order in the cartesian product. For instance, $\Z\C\C\DD$ refers to a term which has a single point at the bottom, then somewhere above it (and to the left or to the right of it) there is an element from $\C$, then another element of $\C$ is further above the previous one, and above all of this there is an element from $\DD$. Schematically, it could look something like the class in Figure~\ref{fig:order}.

\begin{figure}[ht]
  \centering
  \begin{tikzpicture}
    \draw (-0.5,0.5) rectangle (0.5,1.5) node[pos=0.5]{$\DD$};
    \draw (-3,-0.5) rectangle (-2,0.5) node[pos=0.5]{$\C$};
    \draw (-1.5,-1.5) rectangle (-0.5,-0.5) node[pos=0.5]{$\C$};
    \filldraw[black] (-2,-2) circle (2pt);
    \draw (-1.6,-2.2) node {$\Z$};
  \end{tikzpicture}
  \caption{An example of a class which would correspond to a term $\Z\C\C\DD$ in a combinatorial specification that preserves the vertical (bottom-to-top) order of elements.}
  \label{fig:order}
\end{figure}

Let $\V = \{\Z,\Z^*,\C_0,\ldots,\C_r,\C^*_0,\ldots,\C_r^*\}$ refer to the collection of all variables in polynomials $f_i/f_i^*$. When we do not care to distinguish between $f_i$ and $f_j$ or $f_i$ and $f_i^*$, we simply write $f$ for a polynomial in variables from $\V$. Similarly, when we do not distinguish between two variables in $\V$, we simply write $X \in \V$. As we just mentioned, each $f$ is a polynomial. We refer to its terms by $\T_\ell$ as in $f = \sum_{\ell=0}^N\T_\ell$. Each term $\T_\ell$ is a product of the variables in $\V$ and for a function $f^*$, each $\T_\ell$ contains exactly one starred variable (there is exactly one rightmost point in each term). 

We represent every griddable permutation from a juxtaposition by a unique gridded version of it. We pick the gridded version that maximises the element on the right-hand side (RHS) of the juxtaposition. The following convention makes this concept explicit. See also Figure~\ref{fig:leftmostgridline}.\\

\noindent\textbf{Convention:} \emph{Every gridline is as far left as possible. I.e. if it was any further left, the object on the RHS would not belong to the designated class.}\\

\noindent Because we can encode ``gridline as left as possible'', we can refer to griddable permutations by their gridded representatives.

\begin{figure}[ht]
  \centering
  \begin{tikzpicture}[scale=0.2]
\fill[white]  (0,0) rectangle +(27,27);
%\draw[help lines, line width=0.1pt, gray] (0,0) grid +(27,27);
% LHS
\filldraw[black, fill=red] (13.5,24.5) circle (15pt);
\filldraw[black] (12.5,25.5) circle (10pt);
\filldraw[black] (11.5,18.5) circle (10pt);
\filldraw[black] (10.5,23.5) circle (10pt);
\filldraw[black] (9.5,17.5) circle (10pt);
\filldraw[black] (8.5,12.5) circle (10pt);
\filldraw[black] (7.5,16.5) circle (10pt);
\filldraw[black] (6.5,26.5) circle (10pt);
\filldraw[black] (5.5,11.5) circle (10pt);
\filldraw[black] (4.5,2.5) circle (10pt);
\filldraw[black] (3.5,9.5) circle (10pt);
\filldraw[black] (2.5,1.5) circle (10pt);
\filldraw[black] (1.5,10.5) circle (10pt);
\filldraw[black] (.5,.5) circle (10pt);
% RHS
\filldraw[black] (26.5,22.5) circle (10pt);
\filldraw[black] (25.5,21.5) circle (10pt);
\filldraw[black] (24.5,20.5) circle (10pt);
\filldraw[black] (23.5,19.5) circle (10pt);

\filldraw[black] (22.5,15.5) circle (10pt);
\filldraw[black] (21.5,14.5) circle (10pt);
\filldraw[black] (20.5,13.5) circle (10pt);

\filldraw[black] (19.5,8.5) circle (10pt);
\filldraw[black] (18.5,7.5) circle (10pt);
\filldraw[black] (17.5,6.5) circle (10pt);
\filldraw[black] (16.5,5.5) circle (10pt);
\filldraw[black] (15.5,4.5) circle (10pt);
\filldraw[black,fill=red] (14.5,3.5) circle (15pt);
% gridline
\draw[black, line width=0.5mm] (14,-0.5) -- (14,27.5);
\end{tikzpicture}
\caption{\small On the LHS is a permutation from $\C$ while on the RHS is a monotone increasing permutation. The gridline is as far left as possible: if it were shifted further to the left, the red points would form a copy of 21 on the RHS. }
\label{fig:leftmostgridline}
\end{figure}

Further remarks about the way we represent permutations in a juxtaposition. Let $x,y$ be two vertically consecutive points on the left-hand side of the juxtaposition $\C|\DD$. An object on the RHS (e.g. a sequence of points) is said to be \emph{associated with} $x$ if it is in the horizontal section of the RHS that falls below $x$ and above $y$ on the LHS. If $x$ is the bottom most point on the LHS, then everything below it on the RHS is associated with $x$. See Figure~\ref{fig:xregion} for an example.
\begin{figure}[ht]
  \centering
  \begin{tikzpicture}
    \filldraw[black] (1,2) circle (2pt) node[above left]{$y$};;
    \filldraw[black] (2,4) circle (2pt);
    \filldraw[black] (3,1) circle (2pt);
    \filldraw[black] (4,3) circle (2pt) node[above left]{$x$};
    
    \filldraw[orange, opacity=1] (4.2,0) rectangle (6.3,1);
    \filldraw[orange, opacity=0.66] (4.2,1) rectangle (6.3,2);
    \filldraw[orange, opacity=0.33] (4.2,2) rectangle (6.3,3);
    \filldraw[orange, opacity=0.1] (4.2,3) rectangle (6.3,4);
    \draw[black, line width=0.5mm] (4.1,0) -- (4.1,4.5); % gridline
    \draw[thin] (2.2,4)--(6.5,4);
    \draw[thin] (4.2,3)--(6.5,3);
    \draw[thin] (1.2,2)--(6.5,2);
    \draw[thin] (3.2,1)--(6.5,1);
  \end{tikzpicture}
  \caption{The shaded regions on the RHS each correspond to a gap between two vertically consecutive points on the LHS. The part of RHS associated with $x$ is the second least opaque region.}
  \label{fig:xregion}
\end{figure}

% Furthermore, to be able to refer to the boxes inside inflations/substitutions by their altitude, we adopt the superscript numbering $\C^j = \C_{\sigma^{-1}(j)}$, meaning that $\C^j$ is inflating the $j$th point of $\sigma$ when counting from bottom to top, i.e. the point $(j,\sigma(j))$ in $\sigma$. Sometimes, we arrange all classes that inflate $\sigma$ in the bottom-to-top order rather than left-to-right. This will be denoted by double square bracket instead of the original single square brackets, i.e. $\sigma\vert{\C^1,\ldots,\C^n}$ means that the order of classes inside bracket is vertical, reading bottom to top. For future use, notice that in left-to-right order, $\C_{|\sigma|}$ refers to the class that inflates the rightmost point of $\sigma$. In vertical order, this class is labelled as $\C^{\sigma(|\sigma|)}$.

We are going to need the following operators: $\Omega_0, \Omega_1, \Omega_{10}, \Omega_{01}, \Omega_{11}, \Omega_\infty$. They act on permutation classes, such as $\C$, and represent various forms/stages of juxtaposing a monotone class next to $\C$, in other words, $\Omega_i: \C \mapsto \DD$, where both $\C$ and $\DD$ are context-free permutation classes. It is desirable that $\Omega_i$ respects the rightmost point according to the following rules.
\begin{enumerate}
\item If the operator's code begins with $1$, namely $\Omega_1,\Omega_{10},\Omega_{11}$, then the operator can only be applied to a class with a starred class, or alternatively, to a starless class occuring below the starred class.
\item If the operator's code ends with $0$ or $\infty$, namely $\Omega_0, \Omega_{10}, \Omega_\infty$, then its output is starless. If the operator's code ends with $1$, namely $\Omega_1, \Omega_{01}, \Omega_{11}$, then every term of the output contains exactly one rightmost point (a starred class or point).
\end{enumerate}
Rules 1 and 2 capture the observations that: 1. if we juxtapose a monotone (increasing) class next to any class $\C$, $\C|\M$, then the leftmost/lowest point on the RHS must be below the rightmost point on the LHS. And 2., juxtaposing a class on the right sometimes takes over the rightmost point from the class on the left.

Let $\C = \C_0^*$ admit a specification combinatorially isomorphic to $\rS$. Then $\Omega_0(\C_0^*)$ is an operator which juxtaposes a class (starred or not) with an empty space/sequence on the right. Notice, in particular, that $\Omega_0$ distributes over both $+$ and $\times$, and that it erazes ${}^*$. 

\begin{align*}
  \Omega_0(\C_0^*) &= f_0^*(\Omega_0(\Z),\Omega_0(\C_0),\ldots,\Omega_0(\C_{i_0}^*),\Omega_0(\C_{i_0+1}),\ldots,\Omega_0(\C_r))\\
       &\vdots\\
  \Omega_0(\C_r^*) &= f_r^*(\Omega_0(\Z),\Omega_0(\C_0),\ldots,\Omega_0(\C_{i_r}^*),\Omega_0(\C_{i_r+1}),\ldots,\Omega_0(\C_r))\\
  \Omega_0(\C_0) &= f(\Omega_0(\Z),\Omega_0(\C_0),\ldots,\Omega_0(\C_r))\\
       &\vdots\\
  \Omega_0(\C_r) &= f_r(\Omega_0(\Z),\Omega_0(\C_0),\ldots,\Omega_0(\C_r))\\
  \Omega_0(\Z) &= \Z\\
  \Omega_0(\Z^*) &= \Z
\end{align*}

Similarly, $\Omega_\infty$ is distributive over both operations $+$ and $\times$. It is the operator which juxtaposes a class, whether starred or not, with $\M$ on the right. As $\Omega_0$, it also erazes the ${}^*$. We get the following definition of $\Omega_\infty$.

\begin{align*}
  \Omega_\infty(\C_0^*) &= f_0^*(\Omega_\infty(\Z),\Omega_\infty(\C_0),\ldots,\Omega_\infty(\C_{i_0}^*),\Omega_\infty(\C_{i_0+1}),\ldots,\Omega_\infty(\C_r))\\
       &\vdots\\
  \Omega_\infty(\C_r^*) &= f_r^*(\Omega_\infty(\Z),\Omega_\infty(\C_0),\ldots,\Omega_\infty(\C_{i_r}^*),\Omega_\infty(\C_{i_r+1}),\ldots,\Omega_\infty(\C_r))\\
  \Omega_\infty(\C_0) &= f(\Omega_\infty(\Z),\Omega_\infty(\C_0),\ldots,\Omega_\infty(\C_r))\\
       &\vdots\\
  \Omega_\infty(\C_r) &= f_r(\Omega_\infty(\Z),\Omega_\infty(\C_0),\ldots,\Omega_\infty(\C_r))\\
  \Omega_\infty(\Z) &= \M\Z\\
  \Omega_\infty(\Z^*) &= \M\Z\\
  \M &= \EE + \M\Z\\
%  \M^* &= \M\Z^*
\end{align*}

It was necessary to define an extra class $\M$ in order for the combinatorial specification of $\Omega_\infty(\C^*)$ to remain closed (hence also context-free). Usually, $\M$ would be considered to be within the $\C_0,\ldots,\C_r$, but we kept it separate for clarity. The starred version of $\M$ is not needed yet. Consult Figure~\ref{fig:omega_infty} to visualise the juxtapositions $\Omega_\infty(\Z)$ and $\Omega_\infty(\Z^*)$.

\begin{figure}[ht]
    \centering
    \begin{tikzpicture}
      \filldraw[black] (-5,1) circle (2pt) node[above]{$\Z/\Z^*$};
      \draw[|->] (-4,1)--(-1,1);
      \draw (-2.5,1.3) node{$\Omega_\infty$};
      \filldraw[black] (0,2) circle (2pt);
      \draw (-0.3,2.2) node {$\Z$};
      \draw[-, very thick] (0.2,2.5) -- (0.2,0);
      \draw[dashed] (0.2,2) -- (2,2);
      \draw (0.5,0.5) rectangle (1.5,1.5) node[pos=0.5]{$\M$};
    \end{tikzpicture}
    \caption{Juxtaposing an increasing sequence to the right of a point $\Z^*$ or $\Z$ (rightmost or not) means that this point will not be rightmost anymore.}
    \label{fig:omega_infty}
\end{figure}

As before, $X_i$ are variables from $\V$. The next operator, $\Omega_1$, represents juxtaposition of a single point on the right of a class $\C^*$. To define $\Omega_1$, we need to view every $f$ as a finite sum of terms $\T_\ell$. 
\begin{align*}
\Omega_1\left(\sum_{\ell=1}^N \T_\ell\right) &= \sum_{\ell=1}^N \Omega_1(\T_\ell).
\end{align*}
By definition, $\T_\ell = X_1X_2\cdots X_m$ for some $m$, with all $X_i \in \V$. Without loss of generality, we let $k$ be the index of a starred class, i.e. $X_k^*$. Then
\begin{align*}
    \Omega_1(\Z) &= \Z^*\Z\\
  \Omega_1(\Z^*) &= \Z^*\Z\\
  \Omega_1(\T_\ell) &=
                      \begin{cases}
                        \Omega_1(X_1^*)\Omega_0(X_2\cdots X_m) \quad &\text{if $k = 1$}\\
                        \Omega_1(X_1)\Omega_0(X_2\cdots X_m) + \Omega_0(X_1)\Omega_1(X_2\cdots X_m),\quad &\text{if $k>1$.}
                      \end{cases}
\end{align*}
The base cases ($\Z$ and $\Z^*$) are drawn in Figure~\ref{fig:omega_1}. The recursive step $\T_\ell$ consists of two cases. Either the bottom-most class/point $X_1$ in $\T_\ell$ is starred (i.e. $X_1 = X_1^*$) or not. If $X_1^*$, then $\Omega_1$ must be applied to it (as it must be applied to a starred class/point or a class/point below it) and $\Omega_0$ is applied to the rest of the classes, $\Omega_0(X_2\cdots X_m)$. If the first class (or point) $X_1$ is not starred, then there are two options. Either apply $\Omega_1$ to $X_1$ and $\Omega_0$ to $X_2\cdots X_m$, or apply $\Omega_0$ to $X_1$ and recursively apply $\Omega_1$ to $X_2\cdots X_m$. Defining operators recursively will be useful when we apply them to permutation classes iteratively.
\begin{figure}[ht]
 \centering
    \begin{tikzpicture}
      \filldraw[black] (-5,1.5) circle (2pt) node[above]{$\Z/\Z^*$};
      \draw[|->] (-4,1.5)--(-1,1.5);
      \draw (-2.5,1.8) node{$\Omega_1$};
%      \filldraw[black] (0,2) circle (2pt);

      \filldraw[black] (0,2) circle (2pt);
      \draw (-0.3,2.2) node {$\Z$};
      \draw[-, very thick] (0.2,2.5) -- (0.2,1);
      \draw[dashed] (0.2,2) -- (2,2);
      \filldraw[black] (1.2,1.6) circle (2pt);
      \draw (1.6,1.6) node {$\Z^*$};
%      \draw (0.5,0.5) rectangle (1.5,1.5) node[pos=0.5]{$\M$};
    \end{tikzpicture}
    \caption{Juxtaposing a single point to the right of a point $\Z$ or $\Z^*$ means that the original point is not the rightmost anymore. Also, the new point is the rightmost point.}
    \label{fig:omega_1}
\end{figure}

Having notation in place, we define $\Omega_{10}$ and $\Omega_{11}$ as those are fundamentally different from $\Omega_{01}$. As before, both $\Omega_{10}$ and $\Omega_{11}$ are distributive over $+$. It is therefore sufficient to define each of them on $\T_\ell = X_1\cdots X_m$ for some $m$, and the $k$th element starred. Operator $\Omega_{10}$ represents juxtaposing $\C^*$ on the right with a monotone increasing class that tracks its the bottom-most (also left-most) point. As usual, $\Omega_{10}$ erazes the star.
\begin{align*}
  \Omega_{10}(\Z) &= \Z\M\Z\\
  \Omega_{10}(\Z^*) &= \Z\M\Z\\
  \Omega_{10}(\T_\ell) &=
                      \begin{cases}
                        \Omega_{10}(X_1^*)\Omega_\infty(X_2\cdots X_m) \quad &\text{if $k = 1$}\\
                        \Omega_{10}(X_1)\Omega_\infty(X_2\cdots X_m) + \Omega_0(X_1)\Omega_{10}(X_2\cdots X_m),\quad &\text{if $k>1$.}
                      \end{cases}
\end{align*}
The base cases (lines 1 and 2 above) are described in Figures~\ref{fig:omega_10}. 
\begin{figure}[ht]
    \centering
    \begin{tikzpicture}
      \filldraw[black] (0,2) circle (2pt);
      \draw (-0.3,2.2) node {$\Z$};
      \draw[-, very thick] (0.2,2.5) -- (0.2,0);
      \draw[dashed] (0.2,2) -- (2.5,2);
      % \filldraw[black] (2.2,1.6) circle (2pt);
      % \draw (2.6,1.6) node {$\Z^*$};
      \draw (1,0.5) rectangle (2,1.5) node[pos=0.5]{$\M$};
      \filldraw[black] (0.8,0.3) circle (2pt);
      \draw (1.2,0.2) node {$\Z$};
    \end{tikzpicture}
    \caption{Whether the LHS point was the rightmost or not, after applying $\Omega_{10}$, it is not. Notice that $\M$ can be empty.}
    \label{fig:omega_10}
\end{figure}
Next, we define $\Omega_{11}$. The base cases are visualised in Figure~\ref{fig:omega11}.
\begin{align*}
  \Omega_{11}(\Z) &= \Z\M\Z\Z\\
  \Omega_{11}(\Z^*) &= \Z\M\Z^*\Z\\
  \Omega_{11}(\T_\ell) &=
                      \begin{cases}
                        \Omega_{11}(X_1^*)\Omega_0(X_2\cdots X_m)+\Omega_{10}(X_1^*)\Omega_{01}(X_2\cdots X_m) \quad &\text{if $k = 1$}\\
                        \Omega_{11}(X_1)\Omega_0(X_2\cdots X_m) + \Omega_{10}(X_1)\Omega_{01}(X_2\cdots X_m) + \Omega_0(X_1)\Omega_{11}(X_2\cdots X_m),\quad &\text{if $k>1$.}
                      \end{cases}
\end{align*}

\begin{figure}[ht]
%  \centering
  \begin{subfigure}[t]{0.4\textwidth}
    \centering
    \begin{tikzpicture}
      \filldraw[black] (0,2) circle (2pt);
      \draw (-0.3,2.2) node {$\Z$};
      \draw[-, very thick] (0.2,2.5) -- (0.2,0);
      \draw[dashed] (0.2,2) -- (2.5,2);
      \filldraw[black] (2.2,1.6) circle (2pt);
      \draw (2.6,1.6) node {$\Z$};
      \draw (1,0.5) rectangle (2,1.5) node[pos=0.5]{$\M$};
      \filldraw[black] (0.8,0.3) circle (2pt);
      \draw (1.2,0.2) node {$\Z$};
    \end{tikzpicture}
    \caption{$\Omega_{11}(\Z)$: the rightmost point on the LHS is not tracked and so is irrelevant.}
    \label{fig:omega_11_1}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.4\textwidth}
    \centering
    \begin{tikzpicture}
      \filldraw[black] (0,2) circle (2pt);
      \draw (-0.3,2.2) node {$\Z$};
      \draw[-, very thick] (0.2,2.5) -- (0.2,0);
      \draw[dashed] (0.2,2) -- (2.5,2);
      \filldraw[black] (2.2,1.6) circle (2pt);
      \draw (2.6,1.6) node {$\Z^*$};
      \draw (1,0.5) rectangle (2,1.5) node[pos=0.5]{$\M$};
      \filldraw[black] (0.8,0.3) circle (2pt);
      \draw (1.2,0.2) node {$\Z$};
    \end{tikzpicture}
    \caption{$\Omega_{11}(\Z^*)$: point on the LHS is not rightmost anymore, ``being rightmost'' gets passed to the point on the RHS.}
    \label{fig:omega11_2}
  \end{subfigure}
  \caption{Juxtaposing a monotone sequence with tracked topmost and bottom most points to the right of a point $\Z$ and $\Z^*$ respectively.}
  \label{fig:omega11}
\end{figure}

The remaining operator, $\Omega_{01}$, does not begin with a `1' and so is easier to define.
\begin{align*}
  \Omega_{01}(\Z) &= \M\Z\Z\\
  \Omega_{01}(\Z^*) &= \M\Z^*\Z\\
  \Omega_{01}(\T_\ell) &= \Omega_{01}(X_1)\Omega_0(X_2\cdots X_m)+\Omega_\infty(X_1)\Omega_{01}(X_2\cdots X_m)
\end{align*}













% \newpage
% \subsection{Example: juxtaposing three monotone increasing classes}

% \textcolor{red}{TODO: this is actually INCORRECT. There needs to be a phantom point on the left top of $\C$.}

% Hence, $\C = \M$ and we are looking to prove that $\C|\M|\M$ has an algebraic generating function. Or in other words, that $\Omega_{11}(\Omega_{11}(\C))$ has an algebraic generating function. The combinatorial specification of $\C$ is as follows.
% \begin{align*}
%   \C &= \Z + \Z\C
% \end{align*}

% Therefore, $\Omega_{11}(\C)$ has the combinatorial specification below.
% \begin{align*}
%   \Omega_{11}(\C) &= \Omega_{11}(\Z) + \Omega_{11}(\Z\C)\\
%                   &=  \Omega_{11}(\Z) + \Omega'_{11}(\Z)\C + \Omega'_{10}(\Z)\Omega_{01}(\C) + \Z\Omega_{11}(\C)
% \end{align*}

% Consequently, $\Omega_{11}(\Omega_{11}(\C))$ has the following combinatorial specification.
% \begin{align*}
%   \Omega_{11}(\Omega_{11}(\C)) &= \Omega_{11}\left(\Omega_{11}(\Z) + \Omega'_{11}(\Z)\C + \Omega'_{10}(\Z)\Omega_{01}(\C) + \Z\Omega_{11}(\C)\right)\\
%                                &=\Omega_{11}(\Omega_{11}(\Z)) + \Omega_{11}(\Omega'_{11}(\Z)\C) + \Omega_{11}(\Omega'_{10}(\Z)\Omega_{01}(\C)) + \Omega_{11}(\Z\Omega_{11}(\C))\\
%                                &= \Omega_{11}(\Omega_{11}(\Z)) +\\
%                                &+ \Omega_{11}(\Omega'_{11}(\Z))\C + \Omega_{11}(\Omega'_{11}(\Z))\Omega_{01}(\C) +\\
%                                &+ \Omega'_{11}(\Omega'_{10}(\Z))\Omega_{01}(\C) + \Omega'_{10}(\Z)\Omega_{11}(\Omega_{01}(\C)) + \Omega'_{10}(\Omega_{10}(\Z))\Omega_{01}(\Omega_{01}(\C)) +\\
%                                &+ \Omega'_{11}(\Z)\Omega_{11}(\C) + \Omega'_{10}(\Z)\Omega_{01}(\Omega_{11}(\C)) + \Z\Omega_{11}(\Omega_{11}(\C))
% \end{align*}

% \subsection{Example: $\Av(321|12)$}

% The case of $\C = \Av(321)$ is interesting in that there are infinitely many simple permutations in $\C$. First, we need to come up with a combinatorial specification for $\Av(321)$ which is context-free, bottom-to-top (or top-to-bottom) and keeps track of the rightmost point. One such specification is below, but first let us give the context-free grammar that we will derive the specification from. As usual, let $\C^*$ denote the class $\C$ that contains the right-most point of the underlying class, $\Av(321)$ in this case. We enumerate $\C$ with Dyck paths going from bottom to top, underneath the diagonal ($R$ - right step, $U$ - up step). 
% \begin{align*}
%   \C^* &= \C R\C^*U + \C RU^*\\
%   \C &= \C R\C U + \EE
% \end{align*}
% Then the combinatorial specification is as follows.
% \begin{align*}
%   \C^* &= \C\C^*\Z + \C\Z^*\\
%   \C &= \C \C \Z + \EE
% \end{align*}
% Reading bottom-to-top. To enumerate $\Av(321|12)$, we need to find the combinatorial specificatio of $\C + \Omega_1(\C) + \Omega_{11}(\C)$. We will need a subset of the combinatorial specification of the following classes: $$\Omega_\phi(X), \Omega_\EE(X), \Omega_1(X), \Omega_\infty(X), \Omega_{10}(X), \Omega_{01}(X), \Omega_{11}(X),$$ where $X \in \{\C, \C^*, \Z\}$. 

\chapter{Packing}
\label{chap:packing}
%\input{packsmall}

\appendix
%\input{permpack}

\phantomsection
\addcontentsline{toc}{chapter}{Bibliography}
\bibliographystyle{alpha}
\bibliography{thesis} 

\end{document}
\message{ !name(thesis.tex) !offset(-448) }
